Decompositions of congruence subgroups of Chevalley groups

### 2015-10-08 dennis-vaserstein.tex
Some fixes in the introduction

\begin{proof}[Proof of \cref{thm:DennisVaserstein}]
First of all, notice that in the special case $\Phi=\rG_2$ \cref{thm:DennisVaserstein} formally follows from \cref{thm:srRI1} and identities~\ref{rel:h-comm}--\ref{rel:h-w}.
Since $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, 
it suffices to consider only the choices for $\Phi$, $r$ and $s$ listed in the table below.
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \\
\rF_4                                & (4, 1)        & 26             & \text{minimal}       & \rC_3                     & 6              & 3\end{array}\]
 \caption{List of representations involved in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}

>>>>

\begin{proof}[Proof of \cref{thm:DennisVaserstein}]
First of all, notice that in the case $\Phi=\rG_2$ \cref{thm:DennisVaserstein} formally follows from \cref{thm:srRI1} and identities~\ref{rel:h-comm}--\ref{rel:h-w}.
Since $A_{rs} = \E(\Phi, R, I)$ implies $A_{sr} = \E(\Phi, R, I)$, 
it suffices to consider only the following choices for $\Phi$, $r$ and $s$.
\begin{table}[htb]
\[\begin{array}{l @{\qquad} l @{\qquad} c @{\quad} c @{\qquad} c @{\qquad} c @{\qquad} c}
\Phi                                 & (s,r)         & |\Lambda(\pi)| & \text{type of $\pi$} & \text{type of $\Delta_r$} & |\Lambda(\pi')|& |\Gamma|  \\ \hline\vphantom{\Bigl(}
\rA_\ell,\ \ell\geqslant 2           & (1,\ell)      & \ell+1         & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rB_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell+1        & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\     
\rC_\ell,\ \ell\geqslant 2           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 1  \\
\rD_\ell,\ \ell\geqslant 4           & (1,\ell)      & 2\ell          & \text{natural}       & \rA_{\ell-1}              & \ell           & 2  \\ 
\rD_\ell,\ \ell\geqslant 4           & (\ell,\ell-1) & 2^{\ell-1}     & \text{half-spinor}   & \rA_{\ell-1}              & \ell           & \ell-2  \\
\rE_\ell,\ \ell=6,7                  & (\ell,2)      & 27, 56         & \text{minimal}       & \rA_{\ell-1}              & \ell           & 3       \\ 
\rE_\ell,\ \ell=6,7                  & (\ell,1)      & 27, 56         & \text{minimal}       & \rD_{\ell-1}              & 2(\ell-1)      & \ell-1  \\
\rF_4                                & (4, 1)        & 26             & \text{minimal}       & \rC_3                     & 6              & 3\end{array}\]
 \caption{List of representations involved in the proof of \cref{thm:DennisVaserstein}.} \label{table:dv-reps}
\end{table}

### 2015-10-08 paper.tex
Some fixes in the introduction

\title {Decompositions of relative Chevalley groups}
\keywords {Chevalley groups, principal congruence subgroups, stability for $\K_1$, bounded generation, $\SL_2$-factorizations. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z} = \{ z_\alpha(s,\xi) \mid \alpha \in \Phi,\ s\in I,\ \xi \in R \}.$

@@

Our first result is a relative version of Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$ the above theorem boils down to~\cite[Proposition~2.3]{AbeSuzNormalSubgroups}. 
Notice that the special case $I=R$ was treated by the second named author in~\cite{Sm12}.

Our next main result is a relative Dennis---Vaserstein decomposition which holds for a broader class of groups of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$ the distance between which equals $d$.
Assume that $\Phi$ and $I$ satisfy either of the following assumptions:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5, 4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\E(\Phi, R)$ in terms of Tits generators and is, thus, a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{CK83, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. 
Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
%TODO: Now check the coefficients of \rk(\Phi) in the expression below
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\end{enumerate}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{Tavgen93, TavgenThesis}),
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} also with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

@@

Applying \cref{thm:DennisVaserstein} and generalizing (in the special case $\Phi=\rD_\ell$) Nikolov's \cite[Proposition~1]{NikProdDecomp} to arbitrary rings we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}

@@

A root subset $S\subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $\Phi=S \cup -S$ (resp. $S = -S$, resp. $S \cap -S=\varnothing$).
Any parabolic subset $S \subseteq \Phi$ can be decomposed into the disjoint union of its reductive and special part, i.e. 
$S = \Sigma_S \sqcup \Delta_S$, where $\Sigma_S \cap (-\Sigma_S) = \varnothing$, $\Delta_S = -\Delta_S$.

@@

\section{Applications} %\label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
\input{subsystem-factorizations}

>>>>

\title {Decompositions of congruence subgroups of Chevalley groups}
\keywords {Chevalley groups, principal congruence subgroups, stability for $\K_1$, bounded generation, $\SL_2$-factorizations. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see~\cite{Tits76}, \cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z}(\Phi) = \{ z_\alpha(s,\xi) \mid s\in I,\ \xi \in R,\ \alpha \in \Phi\}.$

@@

Our first result is a relative version of Gauss decomposition formulated in terms of the \emph{stable rank} of the ideal $I$ (see section~\ref{sec:stability-conditions} for the definition)
\begin{thm}\label{thm:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley $\E(\Phi, R, I)$ group admits the following decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{thm}
In the special case when $I$ is contained in the Jacobson radical of $R$ the above theorem boils down to~\cite[Proposition~2.3]{AbeSuzNormalSubgroups}. 
The special case $I=R$ was treated by the second named author in~\cite{Sm12}.

The next main result is a relative Dennis---Vaserstein decomposition which holds for a broader class of groups of rings as compared to \cref{thm:srRI1}.
\begin{thm}\label{thm:DennisVaserstein}
Let $\Phi \neq \rE_8$ be a reduced irreducible root system and let $\{ r, s \}$ be a pair of distinct terminal vertices of the Dynkin diagram of $\Phi$ the distance between which equals $d$.
Assume that $\Phi$ and $I$ satisfy either of the following assumptions:
\begin{enumerate}
 \item $\asr(I) \leq d$ for $\Phi=\rF_4$ or $\Phi=\rE_\ell$, $\{r, s\} = \{1, \ell \}$,
 \item $\sr(I) \leq d$, otherwise.
\end{enumerate}
Then $\E(\Phi, R, I) = \EP_r(R, I) \cdot \U(\Sigma^-_r \cap \Sigma^-_s, I) \cdot \EP_s(R, I).$ \end{thm}
Even in the special case $I=R$ the above theorem contains several new cases as compared to~\cite[Theorems~2.5, 4.1]{St78}.
The term ``Dennis---Vaserstein decomposition'' is due to A.~Suslin and M.~Tulenbaev (cf.~\cite[Lemma~2.1]{ST76}).

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\E(\Phi, R)$ in terms of Tits generators and is, thus, a quantitative analogue of \cref{prop:Stepanov}
and a relative analogue of the results of~\cite{CK83, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}.
\begin{thm}\label{thm:width} Let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $R=\mathcal{O}_S$ is a Dedekind ring of arithmetic type posessing a real embedding and $\Phi$ is classical of rank $\geqslant2$, then 
$W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\item If $\sr(I) = 1$ and $\Phi$ is arbitrary irreducible root system then 
\[W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1;\]
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime number $p$, then under the assumption of the Generalized Riemann Hypothesis one has
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\end{enumerate}
\end{thm}
The first assertion of the above theorem is essentially a reformulation of a previously known result of O.~Tavgen (see \cite{Tavgen93, TavgenThesis}),
while for the second and the third assertion we give a direct proof.
Compare \cref{thm:width} with a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} which asserts that 
$W(\SL(n, \mathbb{Z}, I), \mathcal{Z}(\Pi))$ is finite for any ideal $I \leq \mathbb{Z}$ and $n \geq 3$.

@@

Applying \cref{thm:DennisVaserstein} and generalizing Nikolov's \cite[Proposition~1]{NikProdDecomp} in the special case $\Phi=\rD_\ell$ to arbitrary rings we obtain the following result.
\begin{thm}\label{thm:spin-sln-prod}
Assume that $\sr(I) \leqslant 2$. Then the group $\Epin_{2\ell}(R, I)=\E(\rD_\ell, R, I)$ is a product of at most $9$ conjugates of its (regularly embedded) subgroups of type $\rA_{\ell-1}$.
\end{thm}

The rest of the article is organized as follows.
In sections \ref{sec:prelim}--\ref{sec:stability-conditions} we introduce principal notation pertaining to Chevalley groups and stability conditions.
In section \ref{sec:factorizations} we prove our parabolic factorizations and thus obtain Theorems~\ref{thm:srRI1}, \ref{thm:DennisVaserstein} and \ref{thm:SL2width}.
Finally, we treat bounded generation and product decompositions separately in section~\ref{sec:applications} where Theorems~\ref{thm:width} and \ref{thm:spin-sln-prod} are proved.

@@

A proper closed subset of roots $S\subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $\Phi=S \cup -S$ (resp. $S = -S$, resp. $S \cap -S=\varnothing$).
Any parabolic subset $S \subseteq \Phi$ can be decomposed into the disjoint union of its \emph{reductive} and \emph{special} parts, i.e. 
$S = \Sigma_S \sqcup \Delta_S$, where $\Sigma_S \cap (-\Sigma_S) = \varnothing$, $\Delta_S = -\Delta_S$.

@@

\section{Applications} \label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
\input{subsystem-factorizations}
